---
title: "Getting started with data.table"
author: "Kiran Ganji"
date: 2018-06-15
categories: ["Data Science"]
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## data.table outperforms data.frame in atleast 4 different areas


The following are few reasons why one should prefer data.table instead of data.frame

- **Easier** to process large datasets
- **Faster** data manipulations
- **Shorter** syntaxes
- **Better** summary representation

## After this presentation, you won't need data.frame because we are going to...

- Create a data.table using **fread, data.table() and as.data.table**
- Understand the two comma notation of data.table
- Subset the data on rows, columns and both
- Create a new column in data.table using **:=**
- Sort a data.table using **setorder**
- Update values using **:=**
- Delete columns using **:=NULL**
- Remove duplicates using **duplicated**
- Aggregate the data using **list() and by=list()**
- Impute missing values using **:=** and **set**
- Apply functions across the rows using **by=1:nrow()** and **.SDcols**


## I am going to use iris data for demonstration. What a cliche right ;)

We will be using the pre-built dataset, **iris**, to get familiar with various syntaxes of data.table. Lets begin by loading the datasets in R

```{r, echo = TRUE}
library(data.table)

data(iris)
DataDfI <- data.table(iris)
```

The above code just loads the dataset into R enivronment. I will be using the variable **DataDfI** from here on in the presentation to indicate iris data

## Creating a data.table using fread(), data.table() and as.data.table

You can create a data.table from vectors, matrices and csv files. Let's take a look at how this can be achieved

```{r, echo=TRUE}

#Using data.table() function - For data frames and vectors
DataDf <- data.table(iris) #Assume iris is a dataframe
write.csv(DataDf,"Data/iris.csv")

#Using data.table() function - For vectors
ids <- letters 
values <- seq(1,26,1)
DataDf <- data.table(Column = ids, Value = values)

#Using as.data.table() - For matrices
DataDf <- as.data.table(as.matrix(iris))

#Using fread - For csv files
DataDf <- fread("Data/iris.csv")

```


## Understanding the two comma notation of data.table

Unlike data.frame, data.table can take 3 arguments. Lets understand them:

```{r, echo=TRUE}
# DataDf[i,j,k]
```

i - row selection
j - column selection
k - aggregation paramters


**Note:** Unlike data.frame, data.table does not require column names to be given in quotes

```{r, echo=TRUE}
DataDfI[Sepal.Length > 5,,]#Select one row
```


## Subsetting the data on rows

Filter conditions on rows can be supplied as the first arguement of the data.table

Subset rows using one condition
```{r, echo=TRUE}
DataDfI[Sepal.Length > 5,,]
```

Subset rows using two condition
```{r, echo=TRUE}
DataDfI[Sepal.Length > 5 & Sepal.Width > 3.2,,]
```

Subset rows greater than mean
```{r, echo=TRUE}
DataDfI[Sepal.Length > mean(Sepal.Length) ,,]
```

## Subsetting the data on columns

One can select a column in data.table by passing the column name as second argument. The result in doing so will be a vector

```{r, echo=TRUE}
# Select rows with greater than mean
DataDfI[,Sepal.Length,]
```

To retain the data.table structure, the columns should be provided in quotes

```{r, echo=TRUE}
# Select rows with greater than mean
DataDfI[,"Sepal.Length",]
```

To retain data.table structure, multiple columns should be provided as vectors

```{r, echo=TRUE}
# Select rows with greater than mean
DataDfI[,c("Sepal.Length","Sepal.Length"),]
```

## Subsetting based on rows and columns

```{r, echo=TRUE}
# Subsetting rows and columns
DataDfI[Sepal.Length > mean(Sepal.Length),c("Sepal.Length","Sepal.Length"),]
```

## Creating a new column in data.table using **:=**

To generate new columns, we use **:=** syntax. Example shown below.

```{r, echo=TRUE}
DataDfI[,Sepal.Area := Sepal.Length * Sepal.Width,]
DataDfI
```

**Note:** The column creation should be done as part of second argument

One can create multiple columns at one stretch

```{r, echo=TRUE}
DataDfI[,`:=`(Sepal.Area = Sepal.Length * Sepal.Width,
              Petal.Area = Petal.Length * Petal.Width),]
DataDfI
```

**Note:** Do not forget to enclose the **:=** by **`**

## Sorting a data.table using **setorder**

We can use **setorder** synatax for sorting a data table. Example shown below

```{r, echo=TRUE}
# Sort based on the descending order of Petal length
setorder(DataDfI,-Petal.Length)
DataDfI
```

```{r, echo=TRUE}
# Sort based on the ascending order of Sepal length
setorder(DataDfI,Sepal.Length)
DataDfI
```

One can sort based on two columns as well

```{r, echo=TRUE}
# Sort based on the ascending order of Sepal length and then descending order of species
setorder(DataDfI,Sepal.Length,-Species)
DataDfI
```


## Updating values using :=

To update values in data.table, one can use **:=** operator. Let's see an example

```{r, echo=TRUE}
DataDfI[Species == "setosa" & Petal.Length == 1.4,Sepal.Area:=45,]
DataDfI[Species == "setosa" & Petal.Length == 1.4,,]
```

One can also use functions to update the specific values

```{r, echo=TRUE}
DataDfI[Species == "setosa" & Petal.Length == 1.4,Sepal.Area:=(Sepal.Length/Sepal.Width),]
DataDfI[Species == "setosa" & Petal.Length == 1.4,,]
```

## Deleting columns using := NULL

To delete columns in data.table, one can use **:= NULL** syntax. Let's see an example

```{r, echo=TRUE}
DataDfI[,Sepal.Area:=NULL,]
DataDfI
```

We can delete multiple columns as well, but the column names should be given as a vector

```{r, echo=TRUE}
DataDfI[,Sepal.Area:=(Sepal.Length*Sepal.Width),]
DataDfI[,c("Sepal.Area","Petal.Area"):=NULL,]
DataDfI
```

## Removing duplicates using duplicated


We can use **duplicated** syntax for removing duplicates from the data.table

```{r, echo=TRUE}
DataDfI <- rbind(DataDfI, DataDfI)
nrow(DataDfI)
```


```{r, echo=TRUE}
DataDfI <- DataDfI[!duplicated(DataDfI),,]
DataDfI
```


## Aggregate the data using list() and by=list()

Aggregating data is easy to execute in data.table. To aggregate the data, we follow two steps:

- Step 1: Define aggregation methods in second argument of data.table and enclose them in list()
- Step 2: Define the aggregation level and pass it as a third argument in data.table and enclose in by = list()


Lets see what is the average Petal.Length in for each Species

```{r, echo=TRUE}
DataDfI[,list(AvgPetalLength = mean(Petal.Length)),by=list(Species)]
```

When aggregating multiple columns one should separate them by commas

Lets see what is the average Petal.Length and Petal.Width in for each Species
colnames(DataDfI)
```{r, echo=TRUE}
DataDfI[,list(AvgPetalLength = mean(Petal.Length),
              AvgPetalWidth = mean(Petal.Width)),
        by=list(Species)]
```
Instead of mean, one can also use functions like median(), sum() etc.

To get count of the rows, one should use **.N**

```{r, echo=TRUE}
DataDfI[,list(Count = .N),by=list(Species)]
```

One can also get the count of uniques for each columns. To obtain this we have to use **length(unique(Columnname))**

```{r, echo=TRUE}
DataDfI[,list(UniquesPetalLength = length(unique(Petal.Length)),
              UniquesPetalWidth = length(unique(Petal.Width))),by=list(Species)]
```

Note: Make sure that you get only one value after aggregation for each row. If you return multiple values for aggregation, the function will throw an error

## Imputing missing values using := and set

Column wise imputation can done as follows

```{r, echo=TRUE}
DataDfI[is.na(Petal.Length),Petal.Length:=0,]
DataDfI[is.na(Petal.Width),Petal.Length:=0,]
```

One can also impute the values in the whole data.table at once as well. We use the **set** syntax

```{r, echo=TRUE}
for(j in names(DataDfI))
  set(DataDfI,which(is.na(DataDfI)),j,0)
```



## Applying functions across the rows using by=1:nrow() and .SDcols

When applying functions which deal with more than 3 columns and when using other complex functions, one has to use an additional **by=1:nrow(datatablename)** argument

```{r, echo=TRUE}
DataDfI[,Species:=as.character(Species),]
DataDfI[,SpeciesLength := nchar(Species),by=1:nrow(DataDfI)]
```


While applying functions only on a subet of columns, use **.SDcols** syntax

```{r, echo=TRUE}
DataDfI[,Sums:=rowSums(.SD, na.rm = T),.SDcols = c("Petal.Length","Petal.Width")]
DataDfI
```

## Summarizing

We did the following in data.table so far:

- Created a data.table using **fread, data.table() and as.data.table**
- Understood the two comma notation of data.table
- Subsetted the data on rows, columns and both
- Created a new column in data.table using **:=**
- Sorted a data.table using **setorder**
- Updated values using **:=**
- Deleted columns using **:=NULL**
- Removed duplicates using **duplicated**
- Aggregated the data using **list() and by=list()**
- Imputed missing values using **:=** and **set**
- Applied functions across the rows using **by=1:nrow()** and **.SDcols**

## Time to forget data.frame because data.table...

- Can process large datasets with **Ease**
- Can execute data manipulations atleast **1.5 times Faster**
- Has **Shorter** syntaxes: Did you observe any code which was more than a line ;)
- Has **Better** summary representation: Bye bye head() and tail()

**Note:** Have a good day and keep data.tabling :D
















